---
title: "ggplot session"
author: "Ilaria Torre"
date: "30 June 2016"
fontsize: 10pt
output: 
  
  beamer_presentation: 
    fig_height: 2
    fig_width: 4
---

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(echo = TRUE, dev = 'pdf', width = 40, breaklines = T, tidy.opts=list(width.cutoff=40))

```

## Sources

- ggplot documentation: <http://docs.ggplot2.org/current/>
- easier tutorials: <http://www.noamross.net/blog/2012/10/5/ggplot-introduction.html>, 
- R Markdown (the tool I used to make these slides), also an R package
- qplot: <http://www.statmethods.net/advgraphs/ggplot2.html>

## Installation

```{r, echo = T, tidy = T}
# install.packages("ggplot2")
library(ggplot2)
# View(iris)
# ggthemr_reset()
```

- Pros: based on grammar, consistent, makes sense, powerful, active community
- Cons: more difficult to start with, computationally intensive (can solve this with qplot)

## Layers

Plots are built as a series of layers, e.g.:
- One layer for the square that will contain the plot
- One layer for points
- One layer for lines
...

Great introduction: <http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html>

## Iris data set

- 150 rows
- 4 measurements
- 1 factor (species) with 3 levels

```{r, echo = T, tidy = T}
View(iris)
```


## Scatterplot

```{r, echo = T, tidy = T}
ggplot()
```

Nothing is happening: ggplot is just the first layer

## Scatterplot

```{r, echo = T, tidy = T}
ggplot() + 
  geom_point(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
```

## Scatterplot

The above is the same as:

```{r, echo = T, tidy = T}
g <- ggplot()
```

----

```{r, echo=TRUE, tidy = T}
g + 
  geom_point(data = iris, aes(x = Sepal.Length, y = Sepal.Width))
```

## Scatterplot

Adding trendline

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth()
```

## Scatterplot

Regression line

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = lm)
```

## Scatterplot

Get rid of the error

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F)
```

## Plot using stat function

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
  stat_summary(geom = "point", fun.y = mean)
```

Be careful where you put data and aes: only if all layers have the same data, x and y, then you can put these inside ggplot()

## Scatterplot

Adding the grouping factor with colour

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + 
  stat_summary(geom = "point", fun.y = mean)
```

## Scatterplot

Adding the grouping factor with shape

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species)) + 
  stat_summary(geom = "point", fun.y = mean)
```

## Scatterplot

Colour AND shape, why not!

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species, color =  Species)) +
  stat_summary(geom = "point", fun.y = mean)
```

## Barplot

``` {r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Species, y = Petal.Width)) + 
  stat_summary(geom = "bar", fun.y = mean)
```

## Barplot

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Species, y = Petal.Width, fill = Species)) + 
  stat_summary(geom = "bar", fun.y = mean)
```

Notice: "fill" instead of "color"

## Error bars

One step back:
- We can plot bars using stat_summary (when we have a full data frame)
- Or as geom_bar (when we have individual points)

```{r, echo = T, tidy = T}
library(stats)
aggregate(iris$Petal.Width, by = list(iris$Species), FUN = "mean")
```

## Error bars

```{r, echo = T, tidy = T}
iris_error <- data.frame(Species = c("setosa", "versicolor", "virginica"), Petal.Width = c(0.246, 1.326, 2.026))
View(iris_error)

g <- ggplot(data = iris_error, aes(x = Species, y = Petal.Width, fill = Species)) + 
  geom_bar(stat = "identity")
g
```

## Error bars

- With a similar data frame + standard error, we can add error bars
- Function "summarySE" can do it for us 
- Source: http://www.cookbook-r.com/Manipulating_data/Summarizing_data/ 

```{r, echo = T, tidy = T}
source('summarySE.R') 

iris_error <- summarySE(iris, measurevar = "Petal.Width", groupvars = "Species")

g <- ggplot(data = iris_error, aes(x = Species, y = Petal.Width, fill = Species)) + 
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = Petal.Width - se, ymax = Petal.Width + se), width = .2)
g
```

## Error bars

Instead of using the summarySE function, you can create a summary data frame with tidyverse. It has the advantage of using the same grammar as ggplot, doesn't require an external library or function, and is faster.

```{r, echo = T, tidy = T, warnings = F, message = F}
# install.packages("tidyverse") 
library(tidyverse)

iris2 <- iris %>% 
	group_by(Species) %>% 
	summarise(
	  N = n(),
		mean = mean(Petal.Width), 
		sd = sd(Petal.Width),
		se = sd / sqrt(50),
		ci = qt(.975, N - 1) * se ) 
```

---

If you want to compare the speed between the summarySE function and the tidyverse approach, run the following code. For small data sets like this it doesn't really matter, if you are working with more data it makes a difference...

```{r, echo = F, fig.cap="speed comparisson between summarySE() and the tidy approach."}
library(microbenchmark)

tidy.sum <- function(df){
  df %>% 
	group_by(Species) %>% 
	summarise(
	  N = n(),
		mean = mean(Petal.Width), 
		sd = sd(Petal.Width),
		se = sd / sqrt(50),
		ci = qt(.975, N - 1) * se )
}

SE.sum <- function(df){
  summarySE(df, measurevar = "Petal.Width", groupvars = "Species")
}

res <- microbenchmark(
  tidy.sum(iris),
  SE.sum(iris)
)
ggplot(res) +
  geom_boxplot(aes(expr, time/1000)) +
  labs(x = "function", y = "time in ms")

```



## Boxplot

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Species, y = Petal.Width, fill = Species)) + 
  stat_boxplot() 
```

## Line plot

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line")
```

## Line plot

Lines and points

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species))
```

## Messing around with parameters

Base plot:
```{r, {r, echo = T, tidy = T}
g <- ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species))
g
```


## Messing around with parameters

- Changing the dimensions of the axes

```{r, echo = T, tidy = T}
g + 
  xlim(0, 10) + 
  ylim(0,10)
```

## Messing around with parameters

- Changing the size of the elements in the graph

```{r, echo = T, tidy = T}
ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length,color = Species)) + 
  stat_summary(fun.y = mean, geom = "line", size = 2) + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species), size = 3)
```

## Messing around with parameters

- Changing the size of the text around the graph (theme layer)

```{r, echo = T, tidy = T}
g + theme(axis.text = element_text(size = 14), 
legend.text = element_text(size = 14), 
legend.title = element_text(size = 14), 
axis.title = element_text(size = 14))

```

## Messing around with the parameters

- Changing the axes names

```{r, echo = T, tidy = T}
g <- g + 
  labs(x = "Sepal Width", y = "Sepal Length")
g
```

## Messing around with the parameters

- Adding a title

```{r, echo = T, tidy = T}
g + 
  labs(x = "Sepal Width", y = "Sepal Length", title = "Iris plot")
```

## Messing around with the parameters

- Adding a title in a different way

```{r, echo = T, tidy = T}
g <- g + 
  ggtitle("Iris plot 2") 
g
```


## Messing around with the parameters

- Custom colours

```{r, echo = T, tidy = T}
coll <- c("violetred", "darkblue", "lightgreen")
```

## Messing around with the parameters

- Custom colours

```{r, echo = T, tidy = T}
g + 
  scale_color_manual(values = coll)
```

## Messing around with the parameters

- Semi Custom colours

```{r, echo = T, tidy = T}
g + 
  scale_color_brewer(palette = "Spectral")
```

Some palette choices here: <http://www.r-bloggers.com/choosing-colour-palettes-part-ii-educated-choices/>

## Messing around with the parameters

- Grey scales

```{r, echo = T, tidy = T}
g <- g + 
  scale_color_grey()
g
```

## Messing around with the parameters

- Change legend title and names

```{r, echo = F, tidy = T}
g <- ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + 
  stat_summary(fun.y = mean, geom = "point", aes(shape = Species)) + 
  labs(x = "Sepal Width", y = "Sepal Length") + 
  ggtitle("Iris plot 2")

g
```

```{r, echo = T, tidy = T}
g + 
  scale_color_grey(name = "New Legend Title", labels = c("se", "ver", "vir"))
```

But this creates 2 different legends!

## Messing around with the parameters

- The reason is that we only specified the "coloured" layer to have a new legend
- We need to specify the "shaped" layer as well!

```{r, echo = T, tidy = T}
g + 
  scale_color_grey(name = "New Legend Title", labels = c("se", "ver", "vir")) +
  scale_shape_discrete(name = "New Legend Title", labels = c("se", "ver", "vir"))
```



```{r, echo = F, tidy = T}
g <- g + 
  scale_color_grey()

g
```

## Messing around with the parameters

- Print-friendly

```{r, echo = T, tidy = T}
g + 
  theme_bw()

```

## Messing around with the parameters

- Delete legend or part of

```{r, echo = T, tidy = T}
g + 
  guides(color = "none", shape = "none")
```

## Additional themes

- Package 'ggthemr': <https://github.com/cttobin/ggthemr>

```{r, echo = T, tidy = T}
# This is not on CRAN yet. You need to install via
# devtools::install_github('cttobin/ggthemr')
library(ggthemr)
ggthemr('fresh')
g
ggthemr_reset()

```

## Additional themes

- Collection of themes, including a replica of Stata graph schemes: <https://github.com/jrnold/ggthemes>
- APA-themed ggplot template: <https://sakaluk.wordpress.com/2015/04/13/2-the-ggplot2-package-your-gateway-drug-to-becoming-an-r-user/>

## Additional themes

Example of the aforementioned APA theme:

```{r, echo = T, tidy = T}
# windowsFonts(Times=windowsFont("TT Times New Roman")) 

apatheme = theme_bw()+
  theme(panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.border=element_blank(),
        axis.line=element_line(),
        text=element_text(family="Times"),
        legend.title=element_blank())

g + apatheme
```

Note: the fond family argument caused an error in my machine; I solved it by running the windowsFonts() function just above. You might not encounter the error though!

## Faceting

```{r, echo = T, tidy = T}
g + 
  facet_wrap(~Species)
```

## Faceting

- 2 grouping variables
- Iris only has one so add another one

```{r, echo = T, tidy = T}
iris2 <- iris
iris$fac <- c("a", "b", "c")

View(iris)
```



## Faceting

```{r, echo = F, tidy = T}
g <- ggplot(data = iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
  stat_summary(fun.y = mean, geom = "line") + stat_summary(fun.y = mean, geom = "point", 
aes(shape = Species)) + 
  theme(axis.text = element_text(size = 14), legend.text = element_text(size = 14), legend.title = element_text(size = 14), axis.title = element_text(size = 14)) + 
  labs(x = "Sepal Width", y = "Sepal Length") + 
  ggtitle("Iris plot 2") + 
  scale_color_grey() + 
  theme_bw()
```

```{r, echo = T, tidy = T}
g + facet_grid(Species~fac)
iris <- iris2
```


## Saving plot 


```{r, echo = T, tidy = T}

ggsave(g, file = "g.pdf")
ggsave(g, file = "g.png")

```
